Speed and strength of epidemic intervention

### 2019-11-14 doc/speedstrength.tex
more work

@@

The study of measles dynamics provides a key example for which the characterization of an epidemic via the reproduction number has been successful.
One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.

Like epidemic strength, epidemic speed measures how disease spreads in the population. 
\swp{Ebola}

\swp{Eaton and Hallett}

@@

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{strength}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of the average individual as a function of time since infection) \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles. We then have:
\begin{equation}
	\hat K(\tau) = (\RR/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread \RR.

@@

\jd{I'm wondering if we should be diagrammatic about these decompositions. I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}

@@

Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation, for example by identifying and isolating infectious individuals.
We then have:
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
\end{equation}
Since $b$ is a distribution (which integrates to 1), the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$. 
We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/figure1.pdf}
\caption{
\textbf{The infection kernel of the HIV.}
A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
B. Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}. The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
C. Increasing the proportion of early transmission reduces the estimate of the reproduction number.
}
\figlab{example}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/figure2.pdf}
\caption{
\textbf{Understanding condom intervention using strength-based decomposition.}
A. Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $L$ due to condom use is constant across the course of infection. 
B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
}
\figlab{condom}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/figure3.pdf}
\caption{
\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
A. Test-and-treat intervention $\tsub{L}{test}$ starts at 1 and saturates after about 5 years.
B. Increasing the amount of early transmission decreases the strength of an epidemic as well as the strength of test-and-treat intervention.
C. Hazard rate $\tsub{h}{test}$ of test-and-treat intervention $\tsub{L}{test}$ peaks and plateaus very quickly and reaches zero after about 5 years.
D. The amount of early transmission has no effect on the speed of an epidemic and small effect on the speed of intervention.
We assume $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau h(\sigma) d\sigma \right)$ and $h(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $12$ months and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 1/6$.
}
\figlab{test}
\end{figure}

>>>>

\swp{r strategy vs K strategy somewhere}

@@

The study of measles dynamics provides a key example for which the characterization of an epidemic via the basic reproduction number has been successfully applied in controlling the disease.
One of the earliest estimates of the measles $\RR_0$ comes from \cite{anderson1982directly} nearly 40 years ago, who gave a range of $14 - 18$.
This, combined with the mean age of infection (4.6 years) and the mean age of vaccination (2.2 years), allowed them to predict that approximately 96\% of the new birth cohort had to be vaccinated to eradicate measles.
Since then, measles control policies have revolved around targetting this critical vaccination threshold \citep{venczel2003measles, wolfson2007has, takahashi2015reduced}.
\swp{More citations later}
\swp{I think I need to end the paragraph more braodly so that it feels like strength vs speed comparison...}

Like epidemic strength, epidemic speed measures how disease spreads in a population.
Despite the prevailing focus on the basic reproduction number in literature \swp{CITE},
using epidemic speed can be sometimes more effective in characterizing the course of an epidemic and evoluating intervention strategies.
Early in an outbreak, the exponential growth rate can be directly measured from incidence data \citep{chowell2003sars, mills2004transmissibility, nishiura2009transmission, nishiura2010pros, ma2014estimating}, even though the basic reproduction number can be difficult to estimate \citep{weitz2015modeling}.
In these scenarios, we expect control efforts that seek to reduce the epidemic speed to be more effective.
The West African Ebola outbreak in 2014 provides an example for this perspective: 
many researchers tried to estimate the basic reproduction number but the major strategies involved slowing down the epidemic by identifying infected individuals and isolating them \citep{pandey2014strategies}.
Strategies that relied on reducing the epidemic strength (e.g., Ebola vaccine trials) were not implemented until much later and had relatively small effect on controlling the outbreak.
\swp{More citations later}

\swp{Eaton and Hallett?}

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/constant_intervention.pdf}
\caption{
\textbf{Effects of constant strength and speed intervention on infection kernels.}
Ebola-like gamma infection kernel $K(\tau)$ (mean: 16.2 days, CV: 0.58, and $\RR_0$: 1.5) is shown in black \citep{park2019practical}.
The infection kernel after applying each intervention strategy $\hat K(\tau)$ is shown in red.
(A) A constant strength intervention with $\theta = 1.5$ is applied to Ebola-like infection kernel.
A constant strength intervention reduces the density by a constant proportion: $\hat K(\tau) = K(\tau)/\theta$; the resulting distribution is equivalent to the intrinsic generation-intervla distribution ($\hat K(\tau) = g(\tau)$) when the strength of intervention matches the strength of epidemic ($\theta = \mathcal R$).
(B) A constant speed intervention with $\phi \approx 0.0267\,\mathrm{days}^{-1}$ is applied to Ebola-like infection kernel.
A constant speed intervention reduces the density exponentially: $\hat K(\tau) = K(\tau) \exp(-\phi \tau)$; the resulting distribution is equivalent to the initial backward generation-interval distribution ($\hat K(\tau) = b(\tau)$) when the speed of intervention matches the speed of epidemic ($\phi = r$).
}
\figlab{constant}
\end{figure}

Assuming that the infection kernel $K$ doesn't change with time, we write:
\begin{equation}
	K(\tau) = \RR g(\tau),
	\eqlab{strength}
\end{equation}
where $g(\tau)$ is the ``intrinsic'' generation-interval distribution (giving the relative infectiousness of an average individual as a function of time since infection) \citep{champredon2015intrinsic}. 
Since $g$ is a distribution, it integrates to 1, and the reproduction number $\RR$ is thus the integral of $K$.

Imagine a control measure that proportionally reduces $K$, for example by protecting a fixed fraction of susceptibles (\figref{constant}A). We then have:
\begin{equation}
	\hat K(\tau) = (\RR/\theta) g(\tau).
\end{equation}
Since $g$ is a distribution, the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\theta=\RR$. We call $\theta$ the ``strength'' of the intervention; transmission is interrupted when the strength of the intervention $\theta$ is larger than the strength of spread \RR.

@@

\jd{I'm wondering if we should be diagrammatic about these decompositions. I guess you've seen some of the pictures I show during talks about constant-speed and constant-strength interventions and the relationship to generation intervals.}
\swp{I tried}

@@

Now imagine an intervention that reduces transmission at a constant hazard rate $\phi$ across the disease generation (\figref{constant}B), for example by identifying and isolating infectious individuals.
We then have:
\begin{equation}
	\hat K(\tau) = K(\tau)\exp(-\phi\tau) = b(\tau)\exp((r-\phi)\tau)
\end{equation}
Since $b$ is a distribution (which integrates to 1), the reduction needed to prevent invasion (or to eliminate disease)  is exactly $\phi=r$. 
We call $\phi$ the ``speed'' of the intervention; transmission is interrupted when the speed of the intervention is faster than the speed of spread.

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/HIV.pdf}
\caption{
\textbf{The infection kernel of the HIV.}
A. The infection kernel of the HIV is approximated using a sum of two gamma distributions. We assume that the baseline proportion of early transmission is 23\% \citep{hayes2006amplified}.
B. Time series of HIV prevalence in pregnant women in South Africa, 1990 - 2010 \citep{barron2013eliminating}. The initial exponential growth rate of the HIV is estimated by fitting a straight line to log-prevalence (1990 - 1997) by minimizing the sum of squares.
C. Increasing the proportion of early transmission reduces the estimate of the reproduction number.
}
\figlab{example}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/condom.pdf}
\caption{
\textbf{Understanding condom intervention using strength-based decomposition.}
A. Condom use is thought to reduce probability of transmission by a similar factor throughout the course of infection; thus the proportional reduction $L$ due to condom use is constant across the course of infection. 
B. The estimated amount of early transmission affects estimated strength of the epidemic but not of a condom-based intervention.
}
\figlab{condom}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[width=\textwidth]{../figure/test_and_treat.pdf}
\caption{
\textbf{Understanding test-and-treat intervention using strength- and speed-based decomposition.}
A. Test-and-treat intervention $\tsub{L}{test}$ starts at 1 and saturates after about 5 years.
B. Increasing the amount of early transmission decreases the strength of an epidemic as well as the strength of test-and-treat intervention.
C. Hazard rate $\tsub{h}{test}$ of test-and-treat intervention $\tsub{L}{test}$ peaks and plateaus very quickly and reaches zero after about 5 years.
D. The amount of early transmission has no effect on the speed of an epidemic and small effect on the speed of intervention.
We assume $\tsub{L}{test}(\tau) = \exp\left(\int_0^\tau h(\sigma) d\sigma \right)$ and $h(\tau) = \tsub{h}{max} (1 - \exp(- K f(\tau)))$, where $f(\tau)$ is a gamma probability density function with a mean of $12$ months and a shape parameter of 2, $K = 4/\max(f(\tau))$, and $\tsub{h}{max} = 1/6$.
}
\figlab{test}
\end{figure}
